\documentclass[a4paper]{article}
\usepackage{graphicx,amssymb}
\usepackage{authblk}

\textwidth=15cm \hoffset=-1.2cm
\textheight=25cm \voffset=-2cm

\pagestyle{empty}

\date{}

\def\keywords#1{\begin{center}{\bf Keywords}\\{#1}\end{center}}

\def\titulo#1{\title{#1}}
\def\autores#1{\author{#1}}

\begin{document}

\titulo{Title?}

\author[1]{Can Firtina}
\author[1]{Can Alkan}
\affil[1]{\small Dept. of Computer Engineering, Bilkent University, 06800 Ankara, Turkey}
%\renewcommand\Authands{ and }

\maketitle
\thispagestyle{empty}

% The abstract

\begin{abstract}
The recent advancements in next-generation sequencing technologies has increased the demand on producing whole genome sequencing data. Along with the increasing demand, analysing and interpreting the sequencing data in different ways with high accuracy and low cost is the current challenge. Variation discovery from a whole genome has been the main interest for analysing the sequencing data. However, variant discovery at high accuracy is still an unsolved problem for which many algorithms have been introduced. Even though there is still no absolute algorithm to solve this problem, there are several variant callers in the field that are commonly used. In addition to high accuracy problem, these tools are expected to be deterministic in the way that same variation discovery should always be reproducible when sequencing data has been applied several times to a variant caller.

Half of the human genome consists of repeats. Therefore, short fragments maps ambiguously onto half of the human genomes. Alignment algorithms such as Burrows-Wheeler Aligner (BWA) aligns the short reads randomly when BWA may align a short read to several locations(?). We realign the short reads of each sequencing data twice by BWA. Our study aims to reveal if a variant caller produces the same variation discovery results each time same sequencing data is aligned and given as input. Otherwise, we state that the variant caller is not deterministic. In our experiment, we have used GATK HaplotypeCaller, GATK UnifiedGenotyper, freebayes, Platypus, and SAMTools variant callers for SNP discovery. We have used Delly and LUMPY for SV discovery. All variant callers are applied to both low-coverage and high-coverage short reads produced by Illumina from 1000 Genomes Project.  All variant callers produced VCF files from corresponding BAM files.

Our results show that all the variant callers but GATK HaplotypeCaller and GATK UnifiedGenotyper produce same variation discovery when the same individual's sequncing data is given as input. However, GATK HaplotypeCaller and GATK UnifiedGenotyper fail to be deterministic when the same individual's sequencing data is given as input to these variant callers. These results raise question about reproducibility of GATK and may account for both false positives and false negatives(?).
\end{abstract}

\keywords{keywords?} % Write down at least 3 Keywords
\end{document}
